Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-08-25 manuscript.Rnw
Flag to include figures into main document
Quantified overlap with Poorter2015

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals. The database has limited overlap with the recent large compilation of \citet{poorter2015} and differs in that a majority of measurements for large plants are individual rather than stand-based. In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. For each analysis, we used different subsets of the data because not all variables were measured in each study. Sample sizes by PFT are summarised in Table \ref{table:counttable}. We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
    \caption{Overview of the data. (a) Variables were measured on up to \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species. (b) Coverage of the dataset across global climate space. Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space. Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
    \label{fig:figoverview}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant. Sample sizes are listed in Table \ref{table:counttable}}
    \label{fig:figrawdata}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area. (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figlmf}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning. Shown are histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three PFTs. Arrows indicate means by PFT. Colours as in Fig. \ref{fig:figlmf}.}
    \label{fig:fighist}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
     \caption{Leaf and stem traits capture variation in biomass partitioning across species. Individual plant data were averaged by species and study combinations. Lines are linear regressions with 95\% confidence intervals. Both regressions were significant ($P \textless 0.01$). $R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
  \label{fig:figlmfspecies}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning. a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and $M_F/A_S$ within each PFT. Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figclimate}
\end{figure}

>>>>

% Flag to include figures in the main text
\newif\ifincludefigs
\includefigsfalse  % or \includefigstrue

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals. The database has very limited overlap (n = 261, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based estimates converted back to invididuals). In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. For each analysis, we used different subsets of the data because not all variables were measured in each study. Sample sizes by PFT are summarised in Table \ref{table:counttable}. We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
    \fi
    \caption{Overview of the data. (a) Variables were measured on up to \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species. (b) Coverage of the dataset across global climate space. Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space. Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
    \label{fig:figoverview}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \fi
    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant. Sample sizes are listed in Table \ref{table:counttable}}
    \label{fig:figrawdata}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \fi
    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area. (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figlmf}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
    \fi
    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning. Shown are histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three PFTs. Arrows indicate means by PFT. Colours as in Fig. \ref{fig:figlmf}.}
    \label{fig:fighist}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
    \fi
     \caption{Leaf and stem traits capture variation in biomass partitioning across species. Individual plant data were averaged by species and study combinations. Lines are linear regressions with 95\% confidence intervals. Both regressions were significant ($P \textless 0.01$). $R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
  \label{fig:figlmfspecies}
\end{figure}

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
    \fi
    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning. a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and $M_F/A_S$ within each PFT. Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figclimate}
\end{figure}
